---
title: 'DATA 607:  Week 2 Assignment, Connecting to SQL'
author: 'Tom Detzel, Spring 2017'
date: "2/5/2017"
output:
  html_document:
    theme: lumen
    toc: yes
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
	echo = TRUE,
	message = TRUE,
	warning = TRUE
)

library(pander)
panderOptions("digits", 2)
panderOptions("table.style", "rmarkdown")
panderOptions("table.alignment.default", "left")

```

## Introduction
This assignment is a refresher on SQL and an exercise in connecting to a database program using R. In this case, we set up a PostgreSQL database on a free cloud service, ElephantSQL, and used the RPostgreSQL package to make a connnection, fetch data using SQL commands, and prepare, analyze and plot it.

Code for this assignment is based on examples in R and help files for RPostgreSQL and ElephantSQL. The password for the db connection will be blanked out in the .rmd file posted on github; I will supply it in the private Blackboard submission.


### Set up the R workspace  

We'll be using seven R packages in this excercise.  

```{r, eval=F, messages=FALSE, warnings=FALSE}

if (!require('dplyr')) install.packages('dplyr')
if (!require("ggplot2")) install.packages('ggplot2')
if (!require("knitr")) install.packages('knitr')
if (!require("RPostgrSQL")) install.packages('RPostgreSQL')
if (!require("ggthemes")) install.packages('ggthemes')
if (!require("pander")) install.packages('pander')
if (!require("tidyr")) install.packages('tidyr')

```

```{r, echo = FALSE, message=FALSE, warning=FALSE}

library(dplyr)
library(ggplot2)
library(ggthemes)
library(RPostgreSQL)
library(knitr)
library(tidyr)

```

***  

### Connect to postgres database stored on ElephantSQL

*RPostgreSQL* reports that the connection is open and the table we want exists.  

```{r, echo=FALSE, message=FALSE, warnings=FALSE}

# db password
pw <- "SEE BLACKBOARD"

```

```{r, messages=F, warnings=F}

# loads the PostgreSQL driver
drv <- dbDriver("PostgreSQL")

# connect to the postgres db, normal setup
con <- dbConnect(drv, dbname = "hlkofihe",
                 host = "babar.elephantsql.com", port = 5432,
                 user = "hlkofihe", password = pw)

# check for the table
dbExistsTable(con, "reviews")

```

***  

### Pull data into R for processing  

A select query fetches the results from the table "reviews" in the remote database, and we check to be sure the connection is closed.

```{r, messages=F, warnings=F}

# get the data in the ratings table
results <- dbGetQuery(con, "SELECT * FROM reviews")

# close connection
dbDisconnect(con)

# view results
str(results)

```
*** 

### Transform into tidyr data frame  

The data is in a poor format for analysis. We need a new variable "Rater" so we can compare ratings by reviewer. The *tidyr* package function *gather* helps reorganize the data into a usable format.

```{r, messages=F, warnings=F}

# make it a real data frame
results2 <- gather(results, Rater, value=Rating, Julie, Mike, Paul, Beth, Dave, Jane)

knitr::kable(head(results2), caption = "The Transformed Data, Rows 1-6")

```

***  
### Movie rating summary  

We use *dplyr* to group and calculate some simple mean ratings for movies and reviewers. For fun, we compare table output from both *kable* and *pander*.  

```{r}

reslts2 <- tbl_df(results2)

raters <- ungroup(results2) %>%
            group_by(Rater) %>%
            summarise(Reviews = n(), Mean_Rating = round(mean(Rating), 2)) %>%
            arrange(desc(Mean_Rating))

movies <- ungroup(results2) %>%
            group_by(Movie) %>%
            summarise(Reviews = n(), Mean_Rating = round(mean(Rating), 2)) %>%
            arrange(desc(Mean_Rating))

# knitr::kable(raters, caption = "Reviewers: Average Ratings")

pander(raters, caption = "Reviewers: Average Ratings", justify = c("left", "left", "left"), split.cells = c(20, 10, 10))

## knitr::kable(movies, caption = "Movies: Average Ratings")

pander(movies, caption = "Movies: Average Ratings", justify = c("left", "left", "left"), split.cells = c(20, 10, 10))

```


***  

### Movie rating plots  

The plots below show ratings for each movie from each reviwer and the distribution of ratings by reviewer. Female reviewers gave sharply higher ratings.

```{r, messages=F, warnings=F}

# try some plots

# ggplot(results2, aes(x=Rater, fill=Rating)) + geom_density(alpha=.3)

ggplot(results2, aes(x=Movie, y=Rating, colour=Rater)) +
    geom_jitter(size = 3.5, aes(shape = Rater), width = .6, height = .5) +
    labs(title = "Reviewer Ratings Compared") +
    theme_solarized_2()
```

***

```{r, messages=F, warnings=F}

ggplot(results2, aes(x=Rater, y=Rating)) + geom_boxplot(fill="lightblue") +
    guides(fill=FALSE) +
    labs(title = "Reviewer Ratings Distribution") +
    theme_solarized_2()

```

